---
title: "Terminology"
author: "Iñaki Ucar, Bart Smeetss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Terminology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>", 
                      fig.width = 6, fig.height = 4, fig.align = "center")
library(ggplot2)
theme_set(theme_bw())
knitr::read_demo("B-terminology", "simmer")
```

Please find below a short explanation of the most important terminologies used in the docs.

* __Arrival__ 
   An arrival is created by a generator. An arrival can be considered a process, an active entity which has a number of activities associated to it and (in general) a limited lifetime in the system. These activities conform to a trajectory specification.

* __Trajectory__
   When a generator creates an arrival, it couples the arrival to a given trajectory. A trajectory is defined as an interlinkage of activities which together form the arrivals' lifetime in the system. Once an arrival is coupled to the trajectory, it will (in general) start processing the activities in the trajectory in the specified order and, eventually, leave the system.
   
* __Activity__ 
   There are different kinds of activities that allow arrivals to interact with resources, perform custom tasks while spending time in the system, move back and forth through the trajectory dynamically... Currently, the set of available activities consist of `seize`, `release`, `timeout`, `set_attribute`, `rollback` and `branch`. See their respective documentation for details (e.g. `?seize`).

* __Generator__ 
   A generator is a process which creates arrivals with a given interarrival time pattern. These arrivals follow a specified trajectory.

* __Resource__
   A resource is in essence a passive entity. It comprises two parts:
    + __Server__ 
      Conceptually, it represents the resource itself. It has a specified capacity and can be seized and released by an arrival.
    + __Queue__ 
      When an arrival tries to seize a resource (tries to access its server) and it is _busy_, this arrival is appended to the queue (of a given length) if there is room for it. If not, the arrival is rejected and immediately leaves the system.


## From terminology to model

In the simulation model below we create a __trajectory__ which we will call `my trajectory`. This trajectory consists of three consecutive __activities__; a `seize`, `timeout` and `release` event. In the `seize` and `release` events a associated __resource__ type is specified. In this case the resource type is called `operator`. 

```{r create-trajectory}
```


Once the trajectory has been created, the simulation object (a `simmer` instance) is set up. One of the first steps is to actually attach a __resource__ instance called `operator`. Next, a __generator__ instance is attached. This generator, called `my_generator`, has an associated trajectory (`my_trajectory` stored in `t0`) and an associated `dist` function. Once `run()` is called the simulation model will start running. The generator `my_generator` will create an __arrival__ at an interarrival time delivered by the function passed to `dist`.

```{r setup-simmer}
```

For more information, see the other vignettes or the function specific help documentation (e.g. `?simmer`).
