---
title: "Other SimPy Examples"
author: "Iñaki Ucar"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Other SimPy Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, cache = FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>",
                      fig.width = 6, fig.height = 4, fig.align = "center")
library(ggplot2)
theme_set(theme_bw())
```

## Introduction

These examples are adapted from the Python package 'SimPy', [here](https://simpy.readthedocs.io/en/latest/examples). Users familiar with SimPy may find these examples helpful for transitioning to simmer. Some very basic material is not covered. Beginners should first read the vignettes 'Introduction', 'Terminology', 'Advanced Trajectory Usage' and 'The Bank Tutorial'.

## Movie Renege

From [here](https://simpy.readthedocs.io/en/latest/examples/movie_renege.html):

> This example models a movie theater with one ticket counter selling tickets for three movies (next show only). People arrive at random times and try to buy a random number (1-6) of tickets for a random movie. When a movie is sold out, all people waiting to buy a ticket for that movie renege (leave the queue).

> Scenario:
>  A movie theatre has one ticket counter selling tickets for three
>  movies (next show only). When a movie is sold out, all people waiting
>  to buy tickets for that movie renege (leave queue).

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(simmer)

TICKETS <- 50     # Number of tickets per movie
SIM_TIME <- 120   # Simulate until
movies <- c("R Unchained", "Kill Process", "Pulp Implementation")

# setup
set.seed(42)
env <- simmer()

#  A moviegoer tries to by a number of tickets (*num_tickets*) for
#  a certain *movie* in a *theater*.
#
#  If the movie becomes sold out, she leaves the theater. If she gets
#  to the counter, she tries to buy a number of tickets. If not enough
#  tickets are left, she argues with the teller and leaves.
#
#  If at most one ticket is left after the moviegoer bought her
#  tickets, the *sold out* event for this movie is triggered causing
#  all remaining moviegoers to leave.
#
moviegoer <- create_trajectory() %>%
  # select a movie and set reneging condition
  set_attribute("movie", function() sample(3, 1)) %>%
  select(function(attr) movies[attr["movie"]]) %>%
  renege_if(function(attr) paste0(movies[attr["movie"]], " sold out")) %>%
  # check if movie is sold out
  branch(function(attr) get_capacity(env, movies[attr["movie"]]) == 0,
         continue = FALSE,
         # sold out, leave
         create_trajectory() %>% timeout(0)
  ) %>%
  # wait for my turn
  seize("counter", 1) %>%
  # buy tickets
  seize_selected(function() sample(6, 1), continue = FALSE,
                 reject = create_trajectory() %>%
                   # not enough tickets, leave after some discussion
                   timeout(0.5) %>%
                   release("counter", 1)
  ) %>%
  renege_abort() %>%
  branch(function(attr) get_server_count(env, movies[attr["movie"]]) > (TICKETS - 2),
         continue = TRUE,
         # trigger the "sold out" event for the movie
         create_trajectory() %>%
           set_capacity_selected(0) %>%
           send(function(attr) paste0(movies[attr["movie"]], " sold out"))
  ) %>%
  timeout(1) %>%
  release("counter", 1) %>%
  # watch the movie
  wait()

# add movies as resources with capacity TICKETS and no queue
lapply(movies, function(i) add_resource(env, i, TICKETS, 0)) %>% invisible

# add ticket counter with capacity 1 and infinite queue
# add moviegoer generator
# start simulation
env %>%
  add_resource("counter", 1) %>%
  add_generator("moviegoer", moviegoer, function() rexp(1, 1 / 0.5), mon=2) %>%
  run(SIM_TIME)

# analysis/results
sold_time <- env %>% get_mon_resources() %>%
  filter(resource != "counter" & capacity == 0)

reneges <- left_join(
    env %>% get_mon_arrivals() %>% filter(finished == FALSE),
    env %>% get_mon_attributes()
  ) %>%
  rename(resource = value) %>%
  mutate(resource = movies[resource]) %>%
  group_by(resource) %>%
  count()

result <- left_join(sold_time, reneges)

# the output
apply(result, 1, function(i) {
  cat("Movie '", i["resource"], "' sold out ", i["time"], 
      " minutes after ticket counter opening.\n",
      "  Number of people leaving queue when film sold out: ", i["n"], "\n", sep="")
}) %>% invisible
```
