---
title: "The Bank Tutorial: Part I"
author: "Duncan Garmonsway"
description: >
  Learn a classical DES model with simmer.
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{04. The Bank Tutorial: Part I}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, cache = FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>",
                      fig.width = 6, fig.height = 4, fig.align = "center")
```

## Introduction

This tutorial is adapted from a tutorial for the Python 2 package 'SimPy',
[here](https://pythonhosted.org/SimPy/Tutorials/TheBank.html).  Users familiar
with SimPy may find this tutorial helpful for transitioning to `simmer`.

## A single customer

In this tutorial we model a simple bank with customers arriving at random. We
develop the model step-by-step, starting out simply, and producing a running
program at each stage.

A simulation should always be developed to answer a specific question; in these
models we investigate how changing the number of bank servers or tellers might
affect the waiting time for customers.

### A customer arriving at a fixed time

We first model a single customer who arrives at the bank for a visit, looks
around at the decor for a time and then leaves. There is no queueing. First we
will assume his arrival time and the time he spends in the bank are fixed.

The arrival time is fixed at 5, and the time spent in the bank is fixed at 10.
We interpret '5' and '10' as '5 minutes' and '10 minutes'.  The simulation runs
for a maximum of 100 minutes, or until all the customers that are generated
complete their trajectories.

Note where these constants appear in the code below.

```{r}
library(simmer)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  timeout(10) %>%
  log_("I must leave")

bank <-
  simmer("bank") %>%
  add_generator("Customer", customer, at(5))

bank %>% run(until = 100)
bank %>% get_mon_arrivals()
```

The short trace printed out by the `get_mon_arrivals` function shows the result.
The program finishes at simulation time 15 because there are no further events
to be executed. At the end of the visit, the customer has no more actions and no
other objects or customers are active.

### A customer arriving at random

Now we extend the model to allow our customer to arrive at a random simulated
time though we will keep the time in the bank at 10, as before.

The change occurs in the arguments to the `add_generator` function. The function
`rexp` draws from an exponential distribution with the given parameter, which in
this case is `1/5`.  See `?rexp` for more details.  We also seed the random
number generator with 10211 so that the same sequence of random numbers will be
drawn every time the script is run.

```{r}
library(simmer)

set.seed(10212)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  timeout(10) %>%
  log_("I must leave")

bank <-
  simmer("bank") %>%
  add_generator("Customer", customer, at(rexp(1, 1/5)))

bank %>% run(until = 100)
bank %>% get_mon_arrivals()
```

The trace shows that the customer now arrives at time 7.839305. Changing the
seed value would change that time.

### More customers

Our simulation does little so far. To consider a simulation with several
customers we return to the simple deterministic model and add more customers.

The program is almost as easy as the first example (A customer arriving at a
fixed time). The main change is in the `add_generator` function, where we
generate three customers by defining three start times. We also increase the
maximum simulation time to 400 in the `run` function. Observe that we need only
one definition of the customer trajectory, and generate several customers who
follow that trajectory. These customers act quite independently in this model.

Each customer also stays in the bank for a different, but non-random, amount of
time.  This is an unusual case, so it requires an unusual bit of R code.  The
`timeout` function accepts either a constant waiting time, or a function that is
called once per customer and returns a single value (e.g. `rexp(1, 1/10)`).  So
we need to create a function that returns a different, but specific, value each
time it is called. We define a function called `loop` to do this.  The `loop`
function returns another function.  In the example, we store that function under
the name `x`. When called, the function `x` returns one of (in the example) 7,
10, and 20, in sequence, wrapping around when it is called a fourth time.

```{r}
# Function to specify a series of waiting times, that loop around
loop <- function(...) {
    time_diffs <- c(...)
    i <- 0
    function() {
      if (i < length(time_diffs)) {
        i <<- i+1
      } else {
        i <<- 1
      }
      return(time_diffs[i])
    }
  }

x <- loop(10, 7, 20)
x(); x(); x(); x(); x()
```

The technical term for the `loop` function is a 'closure'.  How closures work is
beyond the scope of this vignette; if you wish to learn more, then [Advanced
R](http://adv-r.had.co.nz/Functional-programming.html#closures) by Hadley Wickam
has a good explanation.

When we use `loop` in the `timeout` function, we don't need to assign its output
to a name; it can be an 'anonymous' function. It will be called whenever a
customer is about to wait and needs to know how long to wait for.  Because only
three customers are generated, and their first step is the timeout step, they
are assigned 7, 10, and 20 in order.

Note that this code differs from the SimPy in the order that customers are
defined.  Here, the first customer to be defined is the one who arrives at 2 and
waits for 7.  That is because the arguments to `at()` must be in ascending
order.

```{r}
library(simmer)

# Function to specify a series of waiting times in a loop
loop <- function(...) {
    time_diffs <- c(...)
    i <- 0
    function() {
      if (i < length(time_diffs)) {
        i <<- i+1
      } else {
        i <<- 1
      }
      return(time_diffs[i])
    }
  }

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  timeout(loop(7, 10, 20)) %>%
  log_("I must leave")


bank <-
  simmer("bank") %>%
  add_generator("Customer", customer, at(2, 5, 12))

bank %>% run(until = 400)
bank %>% get_mon_arrivals()
```

Alternatively, we can create three different customer trajectories for the three
waiting times.  This is best done by creating an initial template, and then
modifying the waiting time for each copy.

Note that the order in which the customers are defined does not matter this
time, and we can also name each customer.

```{r}
library(simmer)

# Create a template trajectory
customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  timeout(1) %>% # The timeout of 1 is a placeholder to be overwritten later
  log_("I must leave")

# Create three copies of the template
Klaus <- customer
Tony <- customer
Evelyn <- customer

# Modify the timeout of each copy
Klaus[2] <- timeout(trajectory(), 10)
Tony[2] <- timeout(trajectory(), 7)
Evelyn[2] <- timeout(trajectory(), 20)

# Check that the modifications worked
Klaus
Tony
Evelyn

bank <-
  simmer("bank") %>%
  add_generator("Klaus", Klaus, at(5)) %>%
  add_generator("Tony", Tony, at(2)) %>%
  add_generator("Evelyn", Evelyn, at(12))

bank %>% run(until = 400)
bank %>% get_mon_arrivals()
```

Again, the simulations finish before the 400 specified in the `run` function.

### Many customers

Another change will allow us to have more customers. To make things clearer we
do not use random numbers in this model.

The change is in the `add_generator` function, where we use a convenience
function `from_to` to create a sequence of start times for five customers,
starting at time 0, with an interarrival time of 10 between each customer.  One
idiosyncracy of the syntax is that no arrival is created on the `to` time, so we
give it as 41, one unit after the last arrival to be generated.  Another is that
the interarrival time must be specified as a function, hence we define a
constant function `function() {10}`

```{r}
library(simmer)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  timeout(12) %>%
  log_("I must leave")

bank <-
  simmer("bank") %>%
  add_generator("Customer", customer, from_to(0, 41, function() {10}))

bank %>% run(until = 400)
bank %>% get_mon_arrivals()
```

### Many random customers

We now extend this model to allow arrivals at random. In simulation this is
usually interpreted as meaning that the times between customer arrivals are
distributed as exponential random variates. There is little change in our
program.  The only difference between this and the previous example of a single
customer generated at a random time is that this example generates several
customers at different random times.

The change occurs in the arguments to the `add_generator` function. The function
`rexp` draws from an exponential distribution with the given parameter, which in
this case is `1/10`.  See `?rexp` for more details.  We also seed the random
number generator with 1289 so that the same sequence of random numbers will be
drawn every time the script is run.  The 0 is the time of the first customer,
then four random interarrival times are drawn, and a final -1 stops the
generator.

The reason why we cannot use the `from_to` function here is that we want to
control the number of arrivals that are generated, rather than the end-time of
arrival generation.

```{r}
library(simmer)

set.seed(1289)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  timeout(12) %>%
  log_("I must leave")

bank <-
  simmer("bank") %>%
  add_generator("Customer", customer, function() {c(0, rexp(4, 1/10), -1)})

bank %>% run(until = 400)
bank %>% get_mon_arrivals()
```

## A Service counter

So far, the model has been more like an art gallery, the customers entering,
looking around, and leaving. Now they are going to require service from the bank
clerk. We extend the model to include a service counter that will be modelled as
a 'resource'. The actions of a Resource are simple: a customer requests a unit
of the resource (a clerk).  If one is free, then the customer gets service (and
the unit is no longer available to other customers). If there is no free clerk,
then the customer joins the queue (managed by the resource object) until it is
the customer's turn to be served. As each customer completes service and
releases the unit, the clerk can start serving the next in line.

### One Service counter

The service counter is created with the `add_resource` function. Default
arguments specify that it can serve one customer at a time, and has infinite
queueing capacity.

The `seize` function causes the customer to join the queue at the counter.  If
the queue is empty and the counter is available (not serving any customers),
then the customer claims the counter for itself and moves onto the `timeout`
step.  Otherwise the customer must wait until the counter becomes available.
Behaviour of the customer while in the queue is controlled by the arguments of
the `seize` function, rather than by any other functions.  Once the `timeout`
step is complete, the `release` function causes the customer to make the counter
available to other customers in the queue.

Since the activity trace does not produce the waiting time by default, this is
calculated and appended using the `transform` function.

```{r, message = FALSE}
library(simmer)

set.seed(1234)

bank <- simmer()

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("counter") %>%
  log_(function() {paste("Waited: ", now(bank) - get_attribute(bank, "start_time"))}) %>%
  timeout(12) %>%
  release("counter") %>%
  log_(function() {paste("Finished: ", now(bank))})

bank <-
  simmer("bank") %>%
  add_resource("counter") %>%
  add_generator("Customer", customer, function() {c(0, rexp(4, 1/10), -1)})

bank %>% run(until = 400)
bank %>%
  get_mon_arrivals() %>%
  transform(waiting_time = end_time - start_time - activity_time)
```

Examining the trace we see that the first two customers get instant service but
the others have to wait. We still only have five customers, so we cannot draw
general conclusions.

### A server with a random service time

This is a simple change to the model in that we retain the single service
counter but make the customer service time a random variable. As is traditional
in the study of simple queues we first assume an exponential service time.

Note that the argument to `timeout` must be a function, otherwise it would apply
a constant timeout to every customer.

```{r, message = FALSE}
library(simmer)

set.seed(1269)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("counter") %>%
  log_(function() {paste("Waited: ", now(bank) - get_attribute(bank, "start_time"))}) %>%
  # timeout(rexp(1, 1/12)) would generate a single random time and use it for
  # every arrival, whereas the following line generates a random time for each
  # arrival
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter") %>%
  log_(function() {paste("Finished: ", now(bank))})

bank <-
  simmer("bank") %>%
  add_resource("counter") %>%
  add_generator("Customer", customer, function() {c(0, rexp(4, 1/10), -1)})

bank %>% run(until = 400)
bank %>%
  get_mon_arrivals() %>%
  transform(waiting_time = end_time - start_time - activity_time)
```

This model with random arrivals and exponential service times is an example of
an M/M/1 queue and could rather easily be solved analytically to calculate the
steady-state mean waiting time and other operating characteristics. (But not so
easily solved for its transient behavior.)

## Several Service Counters

When we introduce several counters we must decide on a queue discipline. Are
customers going to make one queue or are they going to form separate queues in
front of each counter? Then there are complications - will they be allowed to
switch lines (jockey)? We first consider a single queue with several counters
and later consider separate isolated queues. We will not look at jockeying.

### Several Counters but a Single Queue

Here we model a bank whose customers arrive randomly and are to be served at a
group of counters, taking a random time for service, where we assume that
waiting customers form a single first-in first-out queue.

The only difference between this model and the single-server model is in the
`add_resource` function, where we have increased the capacity to two so that it
can serve two customers at once.

```{r, message = FALSE}
library(simmer)

set.seed(1269)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("counter") %>%
  log_(function() {paste("Waited: ", now(bank) - get_attribute(bank, "start_time"))}) %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter") %>%
  log_(function() {paste("Finished: ", now(bank))})

bank <-
  simmer("bank") %>%
  add_resource("counter", 2) %>% # Here is the change
  add_generator("Customer", customer, function() {c(0, rexp(4, 1/10), -1)})

bank %>% run(until = 400)
bank %>%
  get_mon_arrivals() %>%
  transform(waiting_time = end_time - start_time - activity_time)
```

The waiting times in this model are much shorter than those for the single
service counter. For example, the waiting time for Customer3 has been reduced
from nearly 17 minutes to less than 9. Again we have too few customers processed
to draw general conclusions.

### Several Counters with individual queues

Each counter is now assumed to have its own queue. The programming is more
complicated because the customer has to decide which queue to join. The obvious
technique is to make each counter a separate resource.

In practice, a customer might join the shortest queue.  We implement this
behaviour by first selecting the shortest queue, using the `select` function.
Then we use `seize_selected` to enter the chosen queue, and later
`release_selected`.

The rest of the program is the same as before.

```{r, message = FALSE}
library(simmer)

set.seed(1014)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  select(c("counter1", "counter2"), policy = "shortest-queue") %>%
  seize_selected() %>%
  log_(function() {paste("Waited: ", now(bank) - get_attribute(bank, "start_time"))}) %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release_selected() %>%
  log_(function() {paste("Finished: ", now(bank))})

bank <-
  simmer("bank") %>%
  add_resource("counter1", 1) %>%
  add_resource("counter2", 1) %>%
  add_generator("Customer", customer, function() {c(0, rexp(4, 1/10), -1)})

bank %>% run(until = 400)
bank %>%
  get_mon_arrivals() %>%
  transform(service_start_time = end_time - activity_time) %>%
  .[order(.$start_time),]
bank %>%
  get_mon_resources() %>%
  .[order(.$time),]
```

The results show that the customers chose the counter with the smallest number.
Unlucky Customer2 who joins the wrong queue has to wait until Customer0 finishes
at time 62.10372, and is the last to leave. There are, however, too few arrivals
in these runs, limited as they are to five customers, to draw any general
conclusions about the relative efficiencies of the two systems.

### The bank with a monitor (aka summary statistics)

We now demonstrate how to calculate average waiting times for our customers.  In
the original SimPy version of this tutorial, this involved using 'Monitors'.  In
simmer, data is returned by the `get_mon_*` family of functions, as has already
been demonstrated.  Here, we simply summarise the data frame returned by the
`get_mon_arrivals` function, using standard R functions.

We also increase the number of customers to 50 (find the number '49' in the
code).

```{r, message = FALSE}
library(simmer)

set.seed(100005)

customer <-
  trajectory("Customer's path") %>%
  seize("counter") %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter")

bank <-
  simmer("bank") %>%
  add_resource("counter", 2) %>%
  add_generator("Customer", customer, function() {c(0, rexp(49, 1/10), -1)})

bank %>% run(until = 1000)

result <-
  bank %>%
  get_mon_arrivals() %>%
  transform(waiting_time = end_time - start_time - activity_time)
```

The average waiting time for 50 customers in this 2-counter system is more
reliable (i.e., less subject to random simulation effects) than the times we
measured before but it is still not sufficiently reliable for real-world
decisions. We should also replicate the runs using different random number
seeds. The result of this run is:

```{r, message = FALSE}
paste("Average wait for ", sum(result$finished), " completions was ",
      mean(result$waiting_time), "minutes.")
```

### Multiple runs

To get a number of independent measurements we must replicate the runs using
different random number seeds. Each replication must be independent of previous
ones, so the environment (bank) must be redefined for each
run, so that the random interarrival times in the `add_generator` function are
generated from scratch.

We take the chunks of code that build the environment (bank) and run the
simulation, and wrap them in the `mclapply` function from the 'parallel'
package.  This function runs each simulation in parallel, using the available
cores in the computer's processor.  Because we use seeds for reproducability, we
pass these to the function that runs the simulation (`function(the_seed)`).

```{r, message = FALSE}
library(simmer)
library(parallel)

customer <-
  trajectory("Customer's path") %>%
  seize("counter") %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter")

mclapply(c(393943, 100005, 777999555, 319999772), function(the_seed) {
  set.seed(the_seed)

  bank <-
    simmer("bank") %>%
    add_resource("counter", 2) %>%
    add_generator("Customer", customer, function() {c(0, rexp(49, 1/10), -1)})

  bank %>% run(until = 400)
  result <-
    bank %>%
    get_mon_arrivals() %>%
    transform(waiting_time = end_time - start_time - activity_time)
  paste("Average wait for ", sum(result$finished), " completions was ",
        mean(result$waiting_time), "minutes.")
}) %>% unlist()
```

The results show some variation. Remember, though, that the system is still only
operating for 50 customers, so the system may not be in steady-state.
